Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2014-07-15 Manuscript/manuscript.tex
Results and discussion edits
Edits to results to switch presentation of metapopulation level results
from outcome to occupancy, and some initial edits to discussion to
reflect those changes.

In the baseline parameterization, pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).  The effect of longevity, however, depended on the shape of the patch quality distribution.  In the simulations with patch quality distributions symmetric around 1, low variance metapopulations favored endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowered the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).  The three alternative parameterizations (models with lattice structure, equal extinction rates, and weak direct infection) produced qualitatively similar results, though pathogen spread in these models was generally more difficult (i.e. fewer pandemics), and the patch quality variance exerted a relatively weaker influence on the epidemiological outcome (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).

\begin{figure}
\centering
\includegraphics[height=5.5cm]{metapop(full).pdf}
\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectious hosts persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome}
\end{figure}

@@

>>>>

In the baseline parameterization, infectious occupancy increased substantially with longevity, while susceptible and total occupancy decreased (Fig~ \ref{poutcome}).  Increasing the variance of the patch quality distribution around a mean of 1 had relatively little effect on average occupancy (Fig~\ref{poutcome}).  However, though the mean showed little change, the variance in both susceptible and infected occupancy increased with the variance of the patch quality distribution, especially at intermediate longevities.  This increase in variance is due to the greater variability in epidemiological outcome in high variance metapopulations, which generate both more pandemics and more disease die-outs than lower variance metapopulations.  The three alternative parameterizations (models with lattice structure, equal extinction rates, and weak direct infection) produced qualitatively similar results, though pathogen spread in these models was generally more limited (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).

\begin{figure}
\centering
\includegraphics[height=5.5cm]{metapop(full).pdf}
\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome}
\end{figure}

@@

The distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.  These impacts are driven largely by the contrasting roles played by low and high quality habitat and their relative balance in the metapopulation.

This work, together with previous studies of disease dynamics in a metapopulation (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}), suggests that there is a general trade-off between forces that facilitate metapopulation stability at the expense of disease spread, and forces that hinder disease spread at the expense of metapopulation stability.  The position of these trade-offs, and their relative importance depend largely on the strength and structure of pathogen transmission.

These previous studies have focused movement as the mechanism behind this trade-off.  Shifting the patch quality distribution towards high quality patches in this model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park.  However, by habitat heterogeneity and metapopulation structure, the model presented here allows us to drill down below the metapopulation level and investigate the role of individual patches.  In the Park model, movement is homogeneous, and increasing movement increases pathogen exposure equally across all patches, whereas in the model here, movement is heterogeneous and directional.  In our model, we're actually holding per capita movement rates constant, and changing the structure and direction of that movement.

Though beyond the scope of the current study, future work could more deeply investigate the effect of changing movement rates across different patch quality distributions.
